---
title: "GC-MS of wood fungi"
author: "Jefferson Pastuña"
date: "2023-08-04"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()
```

## Introduction

The present document aims to record the procedure given for the statistical analysis of secondary metabolites present in the different fungi species. For each step a brief explanation, the code and graphics obtained are included.

The workflow used is taken from the paper ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135). Which offers a wide variety of functions to perform metabolomic profile analysis.

## Before to start

The "notame" package accepts as input a feature table that can be obtained through software such as MZMine, MSDial, among others. In this case, the feature table was obtained with the help of MZmine. The (*.txt) file was slightly modified to obtain the final feature table to notame R package.

Modifications made to the raw (*.txt) file can be summarized in adding and renaming columns. The added columns "Column" and "Ion Mode" allow to analyze samples with different types of columns and with different ionization modes respectively. Also, the cells corresponding to mass and retention time must be renamed so that the package can detect and process it.

## Notame workflow

As a first step for the analysis, all the necessary libraries were installed and loaded.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Loading "notame" library and dependency libraries
library(notame)

# Loading "notame" dependency libraries
install_dependencies

# Loading extra library
library("ComplexHeatmap")
library("readxl")
library("ggplot2")

```

Then, a project path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

ppath <- "../Wood_Fungi/"
init_log(log_file = paste0(ppath, "Result/log.txt"))

```

Next, the feature list was imported.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

data <- read_from_excel(file = "Data/Data_to_notame/MZmine_Feature_List_to_notame.xlsx",
                        sheet = 2, corner_row = 4, corner_column = "K", 
                        split_by = c("Column", "Ion Mode"), rt_limits = c(0, 60))

```

Once the data was read, the next step was to create a MetaboSet in order to create a specific R object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

Finally, each mode was extracted in a single object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

mode <- modes$`Rtx-5MS_EI`

```

As a additional step, we can visualize the raw data in order to inspect the processing routines.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Data extraction of all dataset
test_mode <- modes$`Rtx-5MS_EI`
# Raw boxplot
raw_bx <- plot_sample_boxplots(test_mode,
                               order_by = "QC",
                               fill_by = "QC")
# Raw PCA
raw_pca <- plot_pca(test_mode,
                    center = TRUE,
                    shape = "QC",
                    color = "QC") 
raw_pca + raw_bx

```

## Preprocessing

The first step of the preprocessing is to change the features with value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

mode <- mark_nas(mode, value = 0)

```

Then, features with low detection rate are first flagged and then will be removed. The notame package employs two criteria to select this features. First, is the feature presence in a percentage of QC injections, and then the feature presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

mode <- flag_detection(mode, qc_limit = 0.50, group_limit = 0)

```

With these values, features which that were not detected in the 70% of the QC injections and 90% of sample groups will be discarded.

The next step for preprocessing correspond to drift correction. The drift correction can be applied by smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

corrected <- correct_drift(mode)

```

Then we can visualize the correction for QCs.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

corr_sambx <- plot_sample_boxplots(corrected,
                                   order_by = "QC",
                                   fill_by = "QC")
corr_pca <- plot_pca(corrected,
                     center = TRUE,
                     shape = "QC",
                     color = "QC") 
corr_pca + corr_sambx

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to EI ionization behavior.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

clustered <- cluster_features(corrected, rt_window = 1/60, all_features = TRUE,
                              corr_thresh = 0.95, d_thresh = 0.80)
compressed <- compress_clusters(clustered)

```

We can inspect PCA plot after clustering algorithm.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

comp_sambx <- plot_sample_boxplots(compressed,
                                   order_by = "QC",
                                   fill_by = "QC")
comp_pca <- plot_pca(compressed,
                     center = TRUE,
                     shape = "QC",
                     color = "QC") 
comp_pca + comp_sambx

```

Finally the next step imputes the data using the noise level.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Impute missing values using noise threshold
imputed <- impute_simple(compressed, value = 100, na_limit = 0)

```

We can inspect PCA plot after imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

imp_sambx <- plot_sample_boxplots(imputed,
                                  order_by = "QC",
                                  fill_by = "QC")
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
imp_pca + imp_sambx

```

After data imputation, the data were normalized by PQN.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Probabilistic quotient normalization
pqn_set <- pqn_normalization(imputed,
                             ref = c("qc", "all"),
                             method = c("median", "mean"),
                             all_features = FALSE)

```

We can inspect PCA plot after normalization.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

pqn_sambx <- plot_sample_boxplots(pqn_set,
                                  order_by = "QC",
                                  fill_by = "QC")
pqn_pca <- plot_pca(pqn_set,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
pqn_pca + pqn_sambx

```

# HCA

Droping flagged features and QC sample.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
no_flag <- drop_flagged(pqn_set)
# Drop QC
no_qc <- drop_qcs(no_flag)
# Scaling by autoscaling method
hca_scl <- scale(t(exprs(no_qc)), center = TRUE, scale = TRUE)
hca_scl <- t(hca_scl)
# Adding autoscaling data to notame MetaboSet
scl_set <- no_qc
exprs(scl_set) <- hca_scl
# Extracting feature height table (expression data)
peak_height <- exprs(scl_set)
# Extracting sample information
pdata <- scl_set@phenoData@data
# Extracting feature information
fdata <- scl_set@featureData@data

```

Preparing data and scaling.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Scaling by autoscaling method
hca_scl <- scale(t(exprs(no_qc)), center = TRUE, scale = TRUE)
hca_scl <- t(hca_scl)
# Adding autoscaling data to notame MetaboSet
scl_set <- no_qc
exprs(scl_set) <- hca_scl
# Extracting feature height table (expression data)
peak_height <- exprs(scl_set)
# Extracting sample information
pdata <- scl_set@phenoData@data
# Extracting feature information
fdata <- scl_set@featureData@data
# Adding row and column names
hca_scl <- peak_height
rownames(hca_scl) <- fdata$Feature_ID
colnames(hca_scl) <- pdata$Group

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Dissimilarity matrix
dis <- dist(t(hca_scl), method = "euclidean")
# Hierarchical clustering using Complete Linkage
hca <- hclust(dis, method = "complete" )
# Build dendrogram object from hclust results
dend <- as.dendrogram(hca)
# Extract the data (for rectangular lines)
dend_data <- dendro_data(dend, type = "rectangle")
# Plot the obtained dendrogram
hca_plot <- ggplot(dend_data$segments) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend))+
  geom_text(data = dend_data$labels, aes(x, y, label = label, fontface = 'italic'),
            hjust = 0, size = 3)+
  coord_flip() + scale_y_reverse(limits=c(300,-100), expand = c(0, 0)) +
  theme_dendro()
hca_plot
# Save plot
#ggsave('Result/notame_result/HCA_Plot.pdf', width = 7, height = 4, device = 'pdf', dpi = "print")

```

# UpSet plot of identified metabilites

The dataset before data imputation will be used.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extracting identified metabolite data
upset_set <- compressed[!is.na(compressed@featureData@data$Metabolite),]
# Extract clean data
upset_set <- drop_flagged(upset_set)
# Drop QC
upset_set <- drop_qcs(upset_set)
# Extracting feature height table (expression data)
upset_exprs <- exprs(upset_set)
# Extracting sample information
upset_pdata <- upset_set@phenoData@data
# Extracting feature information
upset_fdata <- upset_set@featureData@data
# Adding row and column names
upset_table <- upset_exprs
rownames(upset_table) <- upset_fdata$Feature_ID
colnames(upset_table) <- upset_pdata$Group

```

Printing the upset plot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Change values to presence (1) and absence (0)
upset_table[upset_table > 0] <- 1
upset_table[is.na(upset_table)] <- 0
# Make the combination matrix
comb_mat = make_comb_mat(upset_table)


# Plot UpSet
png(file="Result/notame_Result/UpSet_pot.png", width = 7, height = 3.5, units = "in", res= 300)
upset_plot <- UpSet(comb_mat,
                    row_names_gp = gpar(fontsize = 10),
                    top_annotation = upset_top_annotation(comb_mat,
                                                          gp = gpar(fill = "#F8766D",
                                                                    col = "#F8766D"),
                                                          add_numbers = TRUE,
                                                          annotation_name_rot = 90,),
                    right_annotation = upset_right_annotation(comb_mat,
                                                              gp = gpar(fill = "#1C6AA8",
                                                                        col = "#1C6AA8"),
                                                              add_numbers = TRUE))
upset_plot
draw(upset_plot)
dev.off()



```











# Metabolite classification

Loadding library for metabite classification.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# ClassyFire package installation
#remotes::install_github('aberHRML/classyfireR')
library(classyfireR)

```

The metabolites were classified using the ClassyFireR.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# InChI key of the metabolites you want to classify
InChI_Keys <- c('2-Heptanone' = "NJICGAVMYWKCMW-GWJXCKMPSA-N")
# Get classification
Classification_List <- purrr::map(InChI_Keys, get_classification)
Classification_List

```




Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

scores <- pca$x %>%                     # Get PC coordinates
  data.frame %>%                        # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%   # Create a new column with the sample names
  left_join(pheno_data )                # Adding metadata
# PCA plot
ggplot(scores, aes(PC1, PC2, shape = QC, color = QC)) +
  geom_point(size = 3) +
  geom_text(label = pheno_data$Group, nudge_x = 0.8, nudge_y = 0.5,show.legend = FALSE,
            cex = 4) +
  guides(x=guide_axis(title = "PC 1 (29.1 %)"), y=guide_axis(title = "PC 2 (14.38 %)")) +
  theme_classic()

# Save plot
#ggsave('Result/GC_MS_EI_Score_Plot.png', width = 5, height = 4, device='png', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
loadings <- EI_pca$rotation %>%           # Extract loadings
  data.frame(Feature_name = rownames(.))  # New column with feat name
```

Creating an artificial table with Feature name and Compound column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
EI_feat_name <- readxl::read_excel("Data/EI_Metabolites.xlsx", 2)

# Creating a new small table of the annotated compounds
EI_compouds_all <- left_join(EI_feat_name, loadings)

# Plotting results
ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.2) +
  theme_classic() + 
  geom_point(data = EI_compouds_all, size = 1) +
  ggrepel::geom_label_repel(data = EI_compouds_all, aes(label = Compound), box.padding = 0.8, label.padding = 0.27, label.r = 0.3, cex = 3) +
  guides(x=guide_axis(title = "PC 1 (22.02 %)"), y=guide_axis(title = "PC 2 (20.39 %)")) +
  ggsci::scale_color_aaas()

# Save plot
#ggsave('Result/GC-MS/GC_MS_EI_Loadings_Plot.png', width = 5, height = 4, device='png', dpi="print")
```

Finish a record.

```{r}

#Loading de mzXML file containing DIA information
mzxml_files_Neg1 <- list.files(pattern = "R2C_neg.mzML")

# Loading libraries of "MSnbase" package
library(MSnbase)

#Getting MS1 information
Data0ondisk_Neg1 <- readMSData(mzxml_files_Neg1, msLevel = 1, mode = "onDisk")

Dataondisk_Neg1<-readMSData(mzxml_files_Neg1, mode = "onDisk")


finish_log()

```
