---
title: "Spectral_Deconvolution"
author: "Jefferson Pastuna"
date: "2024-09-04"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

in progress...

# Before to start

in progress....

# eRah package workflow

eRah package installation and library loading.

```{r echo=TRUE, message=FALSE, warning=FALSE}

# eRah package installation
#install.packages('erah')
# eRah library call
library(erah)

```

Delete unwanted files of data directory and create a directory with the desired files. 

```{r echo=TRUE}

# Delete all file that are not in folders
unlink('Data/Data_to_eRah/*')
# Data folder path
createdt('Data/Data_to_eRah/')

```

Creating a data.frame with chromatograms information.

```{r echo=TRUE}

# Loading (*.CDF) chromatograms
instrumental <- read.csv('Data/Metadata_to_eRah/Metadata_inst.csv')
# Loading metadata of the chromatograms
phenotype <- read.csv('Data/Metadata_to_eRah/Metadata_pheno.csv')
# Merge of metadata information with chromatograms
raw_data <- newExp(instrumental = instrumental,
                   phenotype = phenotype,
                   info = 'Wood fungi')

```

## Compound Deconvolution

Parameters for spectra deconvolution are specified, defining specific criteria on peak width and minimum peak height, noise threshold, and excluding certain ranges of m/z values from processing to improve the accuracy and relevance of chemical analysis.

```{r echo=TRUE}

dec_par <- setDecPar(min.peak.width = 0.7,
                     min.peak.height = 500,
                     noise.threshold = 75,
                     avoid.processing.mz = c(59:64,73:75,147:149),
                     analysis.time = c(3.3,56))

```

To carry out a process in parallel we use the "future" package, which allows us to execute tasks in parallel, improving efficiency and processing speed simultaneously.

```{r echo=TRUE}

plan(future::multisession,
     workers = 14)

```

The deconvolution algorithm is executed using the parameters specified above, the results obtained will be saved in 'MetaboSet' S4 object.

```{r echo=TRUE, warning=FALSE}

dec_data <- deconvolveComp(raw_data,
                           dec_par)

```

## Alignment

Alignment of GC-MS deconvolved compounds with defined parameters.

```{r echo=TRUE}

# Alignment parameters
alig_par <- setAlPar(min.spectra.cor = 0.90,
                     max.time.dist = 4,
                     mz.range = 50:550)
# Alignment
peak_alig <- alignComp(dec_data,
                       alParameters = alig_par)

```

## Missing Compound Recovery

This algorithm allow compound recovery. If a compound appears in at least of four samples than the minimum samples value, then, this compound is searched in the rest of the samples where its concentration has not been registered.


```{r echo=TRUE, warning=FALSE}

peak_find <- recMissComp(peak_alig,
                         min.samples = 2)

```

# Blank subtraction

Exporting feature list to blank subtraction.

```{r echo=TRUE}

# Extracting alignment feature list
feat_list <- alignList(peak_find,
                       by.area = FALSE)
# Exporting alignment feature list
#write.csv(feat_list,
#          file = "Result/eRah_Result/eRah_Feature_List.csv")

```

"notame" package installation for blank subtraction.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame", ref = "v0.3.1")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Loading the last exported eRah feature list (in Excel format according "notame" instruction).

```{r echo=TRUE}

data <- read_from_excel(file = "Data/eRah_Feature_List.xlsx", sheet = 2, 
                        corner_row = 4, corner_column = "G", 
                        split_by = c("Column", "Ion mode"))

```

"notame" MetaboSet construction with the Loading feature list.

```{r echo=TRUE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

Flag contaminant to avoid in subsequent analysis.

```{r echo=TRUE}

# Data extraction
mode <- modes$SHRTX5SilMS_EI
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)
# Removal of contaminants
no_blank <- flag_contaminants(mode,
                              blank_col = "Group",
                              blank_label = "Blank",
                              flag_thresh = 0.39,
                              flag_label = "Contaminant")
# Removal blank group from dataset
no_blank <- no_blank[, no_blank$Group != "Blank"]

# Exporting data to clean identified features
#write_to_excel(no_blank,
#               "Result/eRah_Result/eRah_Feature_List_no_blank.xlsx")

```

# Identification

Text in construction...

```{r echo=TRUE, warning=FALSE}

# Loading NIST 20 (*.msp) library
#nist.database <- importMSP(filename = "E:/NIST_20_Library/Result/NIST20EI_2eRah.MSP",
#                           DB.name = "NIST",
#                           DB.version = "NIST20",
#                           DB.info = "NIST MS Search Export")
# Save library for a posterior faster loading
#save(nist.database, file= "E:/NIST_20_Library/Result/NIST20EI_2eRah.rda")
# Load R library
load("E:/NIST_20_Library/Result/NIST20EI_2eRah.rda")
mslib <- nist.database
# Identification
peak_iden <- identifyComp(peak_find,
                          id.database = mslib,
                          mz.range = NULL,
                          n.putative = 1)
# Identified compounds list
id_list <- idList(peak_iden)
# Exporting identified compounds list
#write.csv(id_list,
#          file = "Result/eRah_Result/NIST_Identification.csv")

```

Exporting spectra data to NIST MS Search identification.

```{r echo=TRUE}

#export2MSP(peak_iden,
#           store.path = "Result/eRah_Result",
#           alg.version = 2)

```

Installation of R package to calculate linear retention index (RI).

```{r echo=TRUE}

# Installation of "MetaboCoreUtils" package
#install.packages("remotes")
#remotes::install_github("rformassspectrometry/MetaboCoreUtils")

# Loading "MetaboCoreUtils" library
library("MetaboCoreUtils")

```

Explain how retention time of n-alkanes was extract?

Read of retention time list of n-alkanes

```{r echo=TRUE}

# Loadding rt of each n-alkane
rti <- data.frame(rtime = c(4.120, 6.196, 8.832, 11.700, 14.574, 17.353,
                            20.001, 22.522, 24.901, 27.167, 29.314, 31.369,
                            33.325, 35.197, 36.991, 38.707, 40.360, 41.950,
                            43.477, 44.952, 46.378, 47.754, 49.093, 50.383,
                            51.666, 53.107),
                  rindex = c(800, 900, 1000, 1100, 1200, 1300, 1400, 1500,
                             1600, 1700, 1800, 1900, 2000, 2100, 2200, 2300,
                             2400, 2500, 2600, 2700, 2800, 2900, 3000,
                             3100, 3200, 3300))

```

## Level 1 identification



## Leval 2 identification

### Tiglic acid

Mirror plot of  alignID number 6

```{r echo=TRUE}

plotSpectra(peak_iden, 6,
            draw.color = "red",
            comp.db = 315495,
            xlim = c(50,120))
# Experimental RI = 941 (Calculated by MS-DIAL software)
# Literature RI = 941 (NIST#: 230359)
# ΔRI = 0

```

Tiglic acid

Mirror plot of  alignID number 7

```{r echo=TRUE}

plotSpectra(peak_iden, 7,
            draw.color = "red",
            comp.db = 315495,
            xlim = c(50,120))
# Experimental RI = 941 (Calculated by MS-DIAL software)
# Literature RI = 941 (NIST#: 230359)
# ΔRI = 0

```

Tiglic acid

Mirror plot of  alignID number 8

```{r echo=TRUE}

plotSpectra(peak_iden, 8,
            draw.color = "red",
            comp.db = 315495,
            xlim = c(50,120))
# Experimental RI = 941 (Calculated by MS-DIAL software)
# Literature RI = 941 (NIST#: 230359)
# ΔRI = 0

```

Tiglic acid

Mirror plot of  alignID number 9

```{r echo=TRUE}

plotSpectra(peak_iden, 9,
            draw.color = "red",
            comp.db = 315495,
            xlim = c(50,120))
# Experimental RI = 941 (Calculated by MS-DIAL software)
# Literature RI = 941 (NIST#: 230359)
# ΔRI = 0

```

### 2-Butenoic acid, 3-methyl-

Mirror plot of  alignID number 10

```{r echo=TRUE}

plotSpectra(peak_iden, 10,
            draw.color = "red",
            comp.db = 315647,
            xlim = c(50,135))
# Experimental RI = 941 (Calculated by MS-DIAL software)
# Literature RI = 947 (NIST#: 1105)
# ΔRI = 6

```




















### Benzoic acid, ethyl ester

Mirror plot of  alignID number 123

```{r echo=TRUE}

# Experimental Retention Index (RI)
indexRtime(c(3.448,3.435,3.534,3.570,3.911,4.050,4.372,4.596,4.682,4.817,
             4.982,5.713,6.160,6.444,6.620,6.849,6.860,7.394,7.657,7.830,
             8.072,8.302,8.409,8.649,9.284,9.432,9.798,9.826,9.898,10.103,
             10.417,10.527,10.694,10.947,11.268,11.614,11.998,12.061,12.160,
             12.695,12.830,13.163,13.321,13.451,13.601,13.920,13.905,14.343,
             14.452,14.574,14.867,14.931,16.015,16.308,16.814,17.095,17.148,
             17.278,17.375,17.503,17.594,17.774,17.863,17.922,18.018,18.077,
             18.619,18.679,18.813,18.891,19.013,19.038,19.114,19.229,19.565,
             19.699,19.752,20.095,20.191,20.305,20.388,20.784,21.005,21.016,
             21.140,21.207,21.303,21.496,21.864,21.958,22.056,22.303,22.483,
             22.551,22.698,22.895,23.232,23.691,23.947,24.104,24.558,24.672,
             25.152,25.330,26.052,26.459,26.388,26.841,27.431,27.842,28.148,
             28.231,28.408,28.445,28.875,29.060,29.780,29.894,30.064,30.205,
             30.483,30.583,30.820,31.036,31.159,31.270,31.362,31.999,32.281,
             32.417,32.546,32.764,32.769,32.939,33.158,33.193,33.408,33.415,
             33.679,33.751,33.831,34.019,34.114,34.482,34.460,34.580,34.808,
             34.936,35.118,35.227,35.456,35.691,35.791,35.882,35.943,36.118,
             36.289,36.366,36.591,36.640,36.692,36.871,36.984,37.048,37.189,
             37.750,37.958,38.073,38.200,38.225,38.600,39.445,40.212,40.356,
             40.857,41.261,41.597,41.946,42.096,42.912,42.977,43.254,43.476,
             43.737,44.409,44.849,44.866,44.782,44.952,45.694,46.376,47.754,
             47.751,48.567,48.886,49.009,49.084,49.279,49.456,50.375,51.091,
             51.308,51.521,51.621,51.793,51.989,52.253,53.003,53.515), rti)

```

### Benzoic acid

Mirror plot of  alignID number 124

```{r echo=TRUE}

plotSpectra(peak_iden, 124,
            draw.color = "red",
            comp.db = 236,
            xlim = c(50,160))
# Experimental Retention Index (RI)
indexRtime(9.049, rti)
# Literature RI = 1171 (NIST#: 290514)
# ΔRI = 10

```

### (E)-2,6-Dimethylocta-3,7-diene-2,6-diol

Mirror plot of  alignID number 133

```{r echo=TRUE}

plotSpectra(peak_iden, 133,
            draw.color = "red",
            comp.db = 265239,
            xlim = c(50,140))
# Experimental Retention Index (RI)
indexRtime(9.7654, rti)
# Literature RI = 1191 (NIST#: 412424)
# ΔRI = 1

```

(E)-2,6-Dimethylocta-3,7-diene-2,6-diol

Mirror plot of  alignID number 134

```{r echo=TRUE}

plotSpectra(peak_iden, 134,
            draw.color = "red",
            comp.db = 265239,
            xlim = c(50,140))
# Experimental Retention Index (RI)
indexRtime(9.7659, rti)
# Literature RI = 1191 (NIST#: 412424)
# ΔRI = 1

```

### Catechol

Mirror plot of  alignID number 140

```{r echo=TRUE}

plotSpectra(peak_iden, 140,
            draw.color = "red",
            comp.db = 8672,
            xlim = c(50,130))
# Experimental Retention Index (RI)
indexRtime(9.9673, rti)
# Literature RI = 1209 (NIST#: 227771)
# ΔRI = 11

```

Catechol

Mirror plot of  alignID number 141

```{r echo=TRUE}

plotSpectra(peak_iden, 141,
            draw.color = "red",
            comp.db = 8672,
            xlim = c(50,130))
# Experimental Retention Index (RI)
indexRtime(9.9682, rti)
# Literature RI = 1209 (NIST#: 227771)
# ΔRI = 11

```

### Benzofuran, 2,3-dihydro-

Mirror plot of  alignID number 152

```{r echo=TRUE}

plotSpectra(peak_iden, 152,
            draw.color = "red",
            comp.db = 27705,
            xlim = c(50,160))
# Experimental Retention Index (RI)
indexRtime(10.4657, rti)
# Literature RI = 1224 (NIST#: 477791)
# ΔRI = 6

```

Benzofuran, 2,3-dihydro-

Mirror plot of  alignID number 154

```{r echo=TRUE}

plotSpectra(peak_iden, 154,
            draw.color = "red",
            comp.db = 27705,
            xlim = c(50,160))
# Experimental Retention Index (RI)
indexRtime(10.4718, rti)
# Literature RI = 1224 (NIST#: 477791)
# ΔRI = 6

```

### Phenol, o-amino-

Mirror plot of  alignID number 156

```{r echo=TRUE}

plotSpectra(peak_iden, 156,
            draw.color = "red",
            comp.db = 318455,
            xlim = c(50,160))
# Experimental Retention Index (RI)
indexRtime(10.3802, rti)
# Literature RI = 1215 (NIST#: 290667)
# ΔRI = 0

```

### 1H-Pyrrole-2,5-dione, 3-ethyl-4-methyl-

Mirror plot of  alignID number 178

```{r echo=TRUE}

plotSpectra(peak_iden, 178,
            draw.color = "red",
            comp.db = 61858,
            xlim = c(50,250))
# Experimental Retention Index (RI)
indexRtime(10.8242, rti)
# Literature RI = 1239 (NIST#: 412808)
# ΔRI = 7

```

### Hydroquinone

Mirror plot of  alignID number 228

```{r echo=TRUE}

plotSpectra(peak_iden, 228,
            draw.color = "red",
            comp.db = 8775,
            xlim = c(50,130))
# Experimental Retention Index (RI)
indexRtime(11.8193, rti)
# Literature RI = 1283 (NIST#: 228148)
# ΔRI = 12

```

Hydroquinone

Mirror plot of  alignID number 231

```{r echo=TRUE}

plotSpectra(peak_iden, 231,
            draw.color = "red",
            comp.db = 8775,
            xlim = c(50,130))
# Experimental Retention Index (RI)
indexRtime(11.8256, rti)
# Literature RI = 1283 (NIST#: 228148)
# ΔRI = 12

```

Hydroquinone

Mirror plot of  alignID number 232

```{r echo=TRUE}

plotSpectra(peak_iden, 232,
            draw.color = "red",
            comp.db = 8775,
            xlim = c(50,130))
# Experimental Retention Index (RI)
indexRtime(11.8273, rti)
# Literature RI = 1283 (NIST#: 228148)
# ΔRI = 12

```

### Indole

Mirror plot of  alignID number 238

```{r echo=TRUE}

plotSpectra(peak_iden, 238,
            draw.color = "red",
            comp.db = 319921,
            xlim = c(50,140))
# Experimental Retention Index (RI)
indexRtime(12.4853, rti)
# Literature RI = 1295 (NIST#: 227886)
# ΔRI = 2

```

Indole

Mirror plot of  alignID number 239

```{r echo=TRUE}

plotSpectra(peak_iden, 239,
            draw.color = "red",
            comp.db = 319921,
            xlim = c(50,140))
# Experimental Retention Index (RI)
indexRtime(12.4894, rti)
# Literature RI = 1295 (NIST#: 227886)
# ΔRI = 2

```

Indole

Mirror plot of  alignID number 242

```{r echo=TRUE}

plotSpectra(peak_iden, 242,
            draw.color = "red",
            comp.db = 319922,
            xlim = c(50,140))
# Experimental Retention Index (RI)
indexRtime(12.4859, rti)
# Literature RI = 1295 (NIST#: 227886)
# ΔRI = 2

```

### 2-Methoxy-4-vinylphenol

Mirror plot of  alignID number 252

```{r echo=TRUE}

plotSpectra(peak_iden, 252,
            draw.color = "red",
            comp.db = 327652,
            xlim = c(50,210))
# Experimental Retention Index (RI)
indexRtime(12.9991, rti)
# Literature RI = 1316 (NIST#: 431439)
# ΔRI = 1

```

2-Methoxy-4-vinylphenol

Mirror plot of  alignID number 253

```{r echo=TRUE}

plotSpectra(peak_iden, 253,
            draw.color = "red",
            comp.db = 327652,
            xlim = c(50,210))
# Experimental Retention Index (RI)
indexRtime(13, rti)
# Literature RI = 1316 (NIST#: 431439)
# ΔRI = 1

```

Ethanone, 1-(2-hydroxy-5-methylphenyl)-

Mirror plot of  alignID number 254

```{r echo=TRUE}

plotSpectra(peak_iden, 254,
            draw.color = "red",
            comp.db = 55545,
            xlim = c(50,240))
# Experimental Retention Index (RI)
indexRtime(13.0013, rti)
# Literature RI = 1316 (NIST#: 235301)
# ΔRI = 1

```

2-Methoxy-4-vinylphenol

Mirror plot of  alignID number 255

```{r echo=TRUE}

plotSpectra(peak_iden, 255,
            draw.color = "red",
            comp.db = 327652,
            xlim = c(50,210))
# Experimental Retention Index (RI)
indexRtime(12.9997, rti)
# Literature RI = 1316 (NIST#: 431439)
# ΔRI = 1

```



## Leval 3 identification

## 1,2-Cyclopentanedione

Mirror plot of  alignID number 12

```{r echo=TRUE}

plotSpectra(peak_iden, 12,
            draw.color = "red",
            comp.db = 292442,
            xlim = c(50,125))
# Experimental RI = 951 (Calculated by MS-DIAL software)
# Literature RI = 934 (NIST#: 162182)
# ΔRI = 17

```








